---
title: "Simulating 2020 Using Markets"
output:
  github_document: 
    df_print: tibble
    toc: true
    toc_dept: 2
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, purl=FALSE}
library(knitr)
opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  comment = "#>",
  fig.width = 9,
  fig.height = 5,
  dpi = 300
)
options(width = 99)
set.seed(seed = 05)
```

Election prediction helps party officials, campaign operatives, and journalists interpret
campaigns in a quantitative manner. Uncertainty is key to a useful election prediction.

The forecast model has become a staple of political punditry. Popularized by the data journalist
at FiveThirtyEight, the forecasting model is a statistical tool used to incorporate a number of
quantitative inputs and produce a _probabilistic_ view of all possible outcomes.

Prediction markets can be used to generate similarly probabilistic views of election outcomes by
utilizing the economic forces of price discovery and risk aversion to overcome the ideological
bias of self-interested traders on a binary options exchange.

Can we possibly use these prediction markets to generate a useful probabilistic simulation of the
electoral college? We'll try and use data from the PredictIt exchange and R code to answer this
question.

```{r packages, warning=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load_current_gh("kiernann/campfin")
pacman::p_load(
  tidyverse,
  jsonlite,
  magrittr,
  janitor,
  rvest,
  usmap
)
```

## Market Data

PredictIt hosts markets for most of the competitive battleground states. We can scrape these
markets using their API and the `jsonlite::fromJSON()` function.

```{r scrape_markets}
markets_prices <-
  fromJSON(txt = "https://www.predictit.org/api/marketdata/all/") %>%
  use_series(markets) %>%
  filter(str_detect(name, "Which party will win (.*) in the 2020 presidential election?")) %>%
  unnest(contracts, names_repair = make_clean_names) %>%
  filter(short_name_2 == "Democratic") %>%
  select(state = short_name, price = last_close_price) %>%
  mutate(state = str_extract(state, "[:upper:]{2}")) %>% 
  arrange(price)
```

From this API, we get probability data for `r nrow(markets_prices)` battleground states.

```{r map_markets, echo=FALSE}
brewer_rdbu <- RColorBrewer::brewer.pal(n = 5, name = "RdBu")
usa_map <- map_data("state") %>% mutate(state = abrev_state(region))
states_ec_map <- left_join(x = usa_map, y = markets_prices)
states_ec_map %>%
  ggplot(mapping = aes(x = long, y = lat, group = group)) +
  geom_polygon(color = "black", mapping = aes(fill = price)) +
  coord_quickmap() +
  scale_fill_gradient2(
    low = brewer_rdbu[1],
    high = brewer_rdbu[5],
    mid = brewer_rdbu[3],
    midpoint = 0.50,
    labels = scales::dollar
  ) +
  theme(
    legend.position = c(0.9, 0.35),
    panel.background = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5),
    axis.text  = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank()
  ) +
  ggtitle("2020 Prediction Markets")
```

These states alone aren't enough to simulate the electoral college in 2020. To predict the
remaining states, we have a few options. The easiest route is to simply assume the party which
won in 2016 will win again in 2020. This isn't a terrible idea, as the majority of states rarely
flip, especially not the `r 50-nrow(markets_prices)` states without a prediction market. We can start
from this assumption and improve upon it very easily.

## Past Data

We can scrape the 2016 election results from Wikipedia, where we can find the percentage of the 
popular vote as well as the number of electoral college votes up for grab in each state.

```{r scrape_past}
past_results <-
  read_html("https://en.wikipedia.org/wiki/2016_United_States_presidential_election") %>%
  html_node("table.wikitable:nth-child(1)") %>%
  html_table(fill = TRUE) %>%
  na_if("–") %>%
  as_tibble(.name_repair = "unique") %>%
  select(1, 4, 5, 8) %>%
  slice(-1, -58, -59) %>%
  set_names(c("state", "dem", "dem_votes", "rep_votes")) %>%
  map_dfc(parse_guess) %>% 
  mutate(
    votes = coalesce(dem_votes, rep_votes),
    dem = parse_double(str_remove(dem, "%"))/100,
    state = state %>% 
      str_remove("\\(at-lg\\)") %>% 
      str_remove(",\\s\\d..$") %>% 
      abrev_state()
  ) %>% 
  group_by(state) %>% 
  summarize(
    past = mean(dem),
    votes = sum(votes)
  )
```

```{r 2016_map, echo=FALSE}
states_2016_map <- left_join(x = usa_map, y = past_results)
states_2016_map %>%
  filter(state != "DC") %>% 
  ggplot(mapping = aes(x = long, y = lat, group = group)) +
  geom_polygon(color = "black", mapping = aes(fill = past)) +
  coord_quickmap() +
  scale_fill_gradient2(
    low = brewer_rdbu[1],
    high = brewer_rdbu[5],
    mid = brewer_rdbu[3],
    midpoint = 0.50,
    labels = scales::dollar
  ) +
  theme(
    legend.position = c(0.9, 0.35),
    panel.background = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5),
    axis.text  = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank()
  ) +
  ggtitle("2016 Popular Vote")
```

## Probabilities

Any good election forecast needs to be _probabilistic_. Professional forecasts take this division of
votes (usually from an aggregate of polls) then calculate the probability distribution around that
range with a series of other factors.

For this simulation, we already have probabilities for `r nrow(markets_prices)` states. PredictIt only
hosts markets for the most competitive states. The reality is, the other `r 51-nrow(markets_prices)` 
contests are fairly noncompetitive. From the density plot below, we can see how the 2016 popular
vote differed for those states _with_ 2020 markets and those without.

```{r vote_range, echo=FALSE}
past_results %>%
  filter(state != "DC") %>%
  mutate(
    market = if_else(
      condition = state %in% markets_prices$state,
      true = "No Market",
      false = "Market Exists"
    )
  ) %>% 
  ggplot(aes(x = past)) + 
  geom_density(aes(fill = market)) +
  facet_wrap(~market, ncol = 1, scales = "free_y") +
  scale_x_continuous(labels = scales::percent) +
  scale_fill_manual(
    guide = FALSE,
    values = c(
      RColorBrewer::brewer.pal(3, "Dark2")[1],
      RColorBrewer::brewer.pal(3, "Dark2")[2]
    )
  ) +
  labs(
    title = "Popular Vote Difference for Battleground States",
    x = "2016 Popular Vote",
    y = "Number of States"
  ) +
  theme(strip.text = element_text(size = 12))
```

For those states without a market, we need to convert these vote shares to probabilities. In
reality, a even a 5% edge results in a significant advantage and a probability difference much
greater than 5%.

To make this conversion, we simulate many new election using the 2016 result. These simulated
elections are normally distributed around the 2016 results with a standard deviation of 0.5 to
account for the uncertainty that's developed in the last 3 years. Below you can see the results of 1,000 simulated elections in Maryland, which voted
`r scales::percent(past_results$past[past_results$state == "MD"])` for the Democratic candidate in
the last election. The area under the curve past 50% is the _probability_ of a democrat winning
again in the next election.

```{r example_range_md, echo=FALSE}
ex_past <- past_results$past[past_results$state == "MD"]
ggplot(data = NULL, mapping = aes(x = rnorm(n = 1000, mean = ex_past, sd = 0.05))) + 
  geom_density(fill = RColorBrewer::brewer.pal(3, "Dark2")[3]) +
  geom_vline(xintercept = 0.5, size = 1) +
  geom_vline(xintercept = ex_past, linetype = 2, size = 1) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "Maryland 2020 Simulated 1000 Times",
    x = "Popular Vote",
    y = "Simulated Occurrences"
  ) +
  geom_label(aes(x = 0.5, y = 4, label = "Winning Threshold")) +
  geom_label(aes(x = ex_past, y = 5, label = "2016 Result")) +
  theme(
    plot.title = element_text(hjust = 0.50),
    axis.text.y.left = element_blank(),
    axis.ticks.y.left = element_blank()
  )
```

Now, lets see the distribution of 1,000 simulated elections in Florida, a much closer election with
only `r scales::percent(past_results$past[past_results$state == "FL"])` of voters supporting the
democratic candidate.

```{r example_range_fl, echo=FALSE}
ex_past <- past_results$past[past_results$state == "FL"]
ggplot(data = NULL, mapping = aes(x = rnorm(n = 1000, mean = ex_past, sd = 0.05))) + 
  geom_density(fill = RColorBrewer::brewer.pal(7, "Dark2")[7]) +
  geom_vline(xintercept = 0.5, size = 1) +
  geom_vline(xintercept = ex_past, linetype = 2, size = 1) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "Florida 2020 Simulated 1000 Times",
    x = "Popular Vote",
    y = "Simulated Occurrences"
  ) +
  geom_label(aes(x = 0.5, y = 4, label = "Winning Threshold")) +
  geom_label(aes(x = ex_past, y = 5, label = "2016 Result")) +
  theme(
    plot.title = element_text(hjust = 0.50),
    axis.text.y.left = element_blank(),
    axis.ticks.y.left = element_blank()
  )
```

We can generate this probability by calculating the average number of simulated elections won by
the democrat. Below, we see how this is done by simulating the Connecticut election 30 times.

```{r example_process}
(ex_past <- past_results$past[past_results$state == "CT"])
(ex_sims <- round(x = rnorm(n = 30, mean = ex_past, sd = 0.05), digits = 4))
(ex_wins <- ex_sims > 0.5)
mean(ex_wins)
```

We can perform such random normal simulations for _every_ state. We'll generate
_10,000_ simulated elections and calculate the percent of those simulated elections where the
democrat won.

```{r simulate_elections}
past_results <- mutate(past_results, prob = NA)
for (i in seq_along(past_results$state)) {
  sims <- rnorm(n = 10000, mean = past_results$past[i], sd = 0.10)
  past_results$prob[i] <- mean(sims > 0.5)
}
```

Below, you can see how the 2016 vote results result in more extreme probabilities.

```{r past_results_hist, echo=FALSE}
past_results %>% 
  mutate(dem = past > 0.5) %>% 
  ggplot(aes(x = past)) +
  geom_histogram(bins = 10, mapping = aes(fill = dem)) +
  geom_vline(xintercept = 0.5, size = 1) +
  labs(
    title = "Popular Vote Distribution",
    x = "Popular Vote",
    y = "2016 States"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(8, "RdBu")[c(1, 8)],
    guide = FALSE
  ) +
  theme(plot.title = element_text(hjust = 0.50))
```

```{r sim_prob_hist, echo=FALSE}
past_results %>% 
  mutate(dem = prob > 0.5) %>% 
  ggplot(aes(x = prob)) +
  geom_histogram(bins = 10, mapping = aes(fill = dem)) +
  geom_vline(xintercept = 0.5) +
  labs(
    title = "Simulated Probability Distribution",
    x = "Popular Vote",
    y = "2016 States"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(8, "RdBu")[c(1, 8)],
    guide = FALSE
  ) +
  theme(plot.title = element_text(hjust = 0.50))
```

This relationship depends entirely on our chosen standard deviation. Again, since we are only
simulating probabilities for those states _without_ markets, these probabilities tend to be extreme.

```{r sim_relationship, echo=FALSE}
past_results %>% 
  filter(state %out% markets_prices$state) %>% 
  ggplot(aes(x = past, y = prob)) +
  geom_hline(yintercept = 0.5, size = 1) +
  geom_vline(xintercept = 0.5, linetype = 2, size = 1) +
  geom_label(aes(label = state, size = votes, fill = prob > 0.5)) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  scale_size_continuous(range = c(2, 10), guide = FALSE) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(8, "RdBu")[c(1, 8)], guide = FALSE) +
  labs(
    title = "Popular Vote Relationship to Simulated Probability",
    x = "2016 Democratic Vote",
    y = "2020 Democratic Probability"
  )
```

## Combine Sources

The efficient market hypothesis holds that our markets are a more accurate method to generate
probabilistic predictions. We will uses these market prices over our simulated elections where we
have them.

```{r join_past_price}
ec <- past_results %>% 
  left_join(markets_prices, by = "state") %>% 
  mutate(
    dem = coalesce(price, prob),
    market = !is.na(price)
  ) %>% 
  select(state, dem, market, votes)
```

```{r sim_relationship_market, echo=FALSE}
past_results %>% 
  left_join(markets_prices, by = "state") %>% 
  mutate(
    dem = coalesce(price, prob),
    market = !is.na(price)
  ) %>% 
  ggplot(aes(x = past, y = dem)) +
  geom_hline(yintercept = 0.5, size = 1) +
  geom_vline(xintercept = 0.5, linetype = 2, size = 1) +
  geom_label(aes(label = state, size = votes, fill = market)) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  scale_size_continuous(range = c(2, 10), guide = FALSE) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(3, "Dark2")[c(1, 2)]) +
  labs(
    title = "Popular Vote Relationship to Simulated Probability",
    fill = "Market Price",
    x = "2016 Democratic Vote",
    y = "2020 Democratic Probability"
  ) +
  theme(legend.position = "bottom")
```

```{r 2020_map, echo=FALSE}
states_2020_map <- left_join(x = usa_map, y = ec)
states_2020_map %>%
  ggplot(mapping = aes(x = long, y = lat, group = group)) +
  geom_polygon(color = "black", mapping = aes(fill = dem)) +
  coord_quickmap() +
  scale_fill_gradient2(
    low = brewer_rdbu[1],
    high = brewer_rdbu[5],
    mid = "white",
    midpoint = 0.50,
    labels = scales::dollar
  ) +
  theme(
    legend.position = c(0.9, 0.35),
    panel.background = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5),
    axis.text  = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank()
  ) +
  ggtitle("2016 Popular Vote")
```

## Electoral College

To simulate the entire electoral college, we simple have to perform the same `sample()` process as
we did with Connecticut, above. To simplify this process, we can create a new `sim_race()` function
that takes a probability and returns a `TRUE` or `FALSE` indicating whether or not the democrat
has won.

```{r}
sim_race <- function(dem = 1-rep, rep = 1-dem) {
  sample(
    size = 1,
    x = c(TRUE, FALSE),
    prob = c(dem, rep)
  )
}
```

With this function, we can simulate every state in the country and count the number of electoral
college votes won by each party.

```{r sim1}
sim1 <- map_lgl(ec$dem, sim_race)
sum(ec$votes[sim1])
```

```{r}
sim1_result <- if_else(
  condition = sum(ec$votes[sim1]) > 269,
  true = "the Democrats did win",
  false = "the Democrats did not win"
)
```

In the above election `r cat(sim1_result)`. To best understand the _range_ of possible outcomes, 
we will perform the same simulation 10,000 times.

```{r}
n <- 10000
sims <- rep(NA, n)
for (i in seq(1, n)) {
  state_outcomes <- map_lgl(ec$dem, sim_race)
  dem_total <- sum(ec$votes[state_outcomes])
  sims[i] <- dem_total
}
```

```{r}
ggplot(data = NULL) +
  geom_histogram(mapping = aes(x = sims, fill = sims > 270)) +
  geom_vline(xintercept = 270, size = 1) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(8, "RdBu")[c(1, 8)], guide = FALSE) +
  labs(
    title = sprintf("%s Simulated Electoral College Outcomes", scales::comma(n)),
    x = "Simulated Democratic Votes",
    y = "Number of Simulations"
  ) 
```

